CODECHECK: An open-science initiative to facilitate sharing of computer programs and results presented in scientific publications.

### 2020-09-17 codecheck-paper.tex
add reference article for 2020-012

\begin{document}
\title{CODECHECK: An open-science initiative to facilitate sharing of computer programs and results presented in scientific publications.}
\author[1,$\ast$]{Daniel N\"{u}st}
\author[2,$\ast$]{Stephen J Eglen}
\affil[1]{Address of author1}
\affil[2]{Address of author2}
\affil[$\ast$]{Contributed equally}
\maketitle
\begin{abstract}
\ldots{}
\end{abstract}

@@

So, assuming that researchers begin to share more artifacts, how might
they be examined to test that they do what they claim? For example, most
scientific journals now require a data sharing statement that outlines
what data the authors have (or will) share. The implementation status of
this varies considerably according to the journal. At one of the
spectrum, there are specialist journals that have been created to accept
``data papers'' (e.g.~\emph{Scientific Data}, \ldots{} NAME COUPLE OF
OTHERS). These journals have established rigorous procedures by which
the data is validated according to standards in each field. At the other
end of the spectrum, we still witness today the infamous statement
``Data available upon reasonable request'' whereby authors,
well-intentioned, often cannot provide the data when readers ask for it.
(ANDREW H: any references to sum up poor/evolving state of data sharing;
Vines et al 2014).

@@

Step 1: the author submit their manuscript along with code and data to
the publishers. The code and data need not be openly available at this
point.

Step 2: the publisher finds a codechecker to check the code. This is
analgous to the publisher finding one or more reviewers to evaluating
the code. However a cruical difference is that we suggest the codechcker
and the author can talk freely to each directly (see next).

Step 3: the codechecker runs the code, based on instructions provided by
the author, and checks if some/all of the results from the paper can be
reproduced. (See later for exact nature of ``reproduced''). If there are
any problems running the code, the codechecker asks the author for help
to resolve the problems. The codechecker then tries to run the code
again. This process iterates until either the codechecker is successful,
or the codechecker concludes the code does not (\emph{what do in this
case}). As part of this process, the codechecker could work entirely
locally on their own compute resource, or in the cloud, e.g.~using
mybinder infrastructure. Using such cloud-based infrastructure allows
for the codechecker and author to collaboratively run the code together.

Step 4: The codechecker writes a certificate stating how the code was
run and includes a copy of outputs (figures, tables) that were
independently generated.

Step 5: the certificate and code/data get deposited on an open archive
(currently Zenodo), and the certificate is given to the PUBLISHER.

Step 6: the Publisher can then use the certificate as part of their
supplementary records for a paper, and give credit to the codechecker
for their work by depositing appropriate records (\emph{bit hazy here}).

\begin{figure}
  \centering
      \includegraphics{/home/stephen/papers/talks/talkfigs/openscience/codecheck_overview.pdf}
  \caption{The detective. Could we number the arrows as steps?
(Implementing the CODECHECK process).  DANIEL: best source for this figure?}
  \label{fig:worfklow}
\end{figure}

@@

This example workflow leaves sufficient room for different
implementations of this process. There are several aspects that can be
considered:

@@

We have a curated list of 15 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons (do we name them all?)
from interesting historical papers, to papers codechecked as part of
peer review, or because of public interest (covid). several covid models
have been done during the time of peer review of these papers, under our
initative rather than requested from journal, but certificate have then
been acknowledged/cited in the paper (LANCET / LSHTM).

@@

@@

What can you expect to find a certificate? Draw an annotated example.

Should we annotate a certificate, at least the cover page, to
demonstrate what we mean by a certificate? We could print a four page
certificate 4-up so that it fits onto one page. Text might be too small
though to read? A 2-up, printed sideways would work well though.

\subsubsection*{Tools}\label{tools}

To run the system, we rely on github to host our work, and use a custom
R pacakge, codecheck, to help with the generation of certificates and
depositing work on zenodo. (Cite zen4R).

You can do all with open infrastructure (e.g.~automated rendering of
register). but you can use your own tools if you prefer.

@@

As a principle, we do not require results to be exactly the same for a
codecheck to pass, simply that the code runs and generates the output
files that the author claims. Stochastic simulations mean that often we
cannot get exactly the same, or even different versions of libraries.
(Cite freesurfer example that generates different outputs on different
operating systems).

\begin{itemize}
\item
  modelDB principle: minimum of one figure should be reproducibe.
  Exampler certificate 2020-016 (I think) which had only a small subset
  of the possible figures. Still useful and worth sharing.
\end{itemize}

>>>>

\begin{document}
\title{CODECHECK: An open-science initiative to facilitate sharing of
  computer programs and results presented in scientific publications.}
\author[1,$\ast$]{Daniel N\"{u}st}
\author[2,$\ast$]{Stephen J Eglen}
\affil[1]{\url{https://orcid.org/0000-0002-0024-5046}}
\affil[2]{\url{https://orcid.org/0000-0001-8607-8025}}
\affil[$\ast$]{Contributed equally}
\maketitle
\begin{abstract}
\ldots{}
\end{abstract}

@@

So, assuming that researchers begin to share more artifacts, how might
they be examined to test that they do what they claim? For example,
most scientific journals now require a data sharing statement that
outlines what data the authors have (or will) share. The
implementation status of this varies considerably according to the
journal. At one of the spectrum, there are specialist journals that
have been created to accept ``data papers'' (e.g.~\emph{Scientific
  Data}, \textit{Geoscience Data Journal}\ldots{} NAME COUPLE OF
OTHERS). These journals have established rigorous procedures by which
the data is validated according to standards in each field. At the
other end of the spectrum, we still witness today the infamous
statement ``Data available upon reasonable request'' whereby authors,
well-intentioned, often cannot provide the data when readers ask for
it.  (ANDREW H: any references to sum up poor/evolving state of data
sharing; Vines et al 2014).

@@

\textbf{Step 1:} the author submit their manuscript along with code and data to
the publishers. The code and data need not be openly available at this
point.

\textbf{Step 2:} the publisher finds a codechecker to check the code. This is
analgous to the publisher finding one or more reviewers to evaluating
the code. However a cruical difference is that we suggest the codechcker
and the author can talk freely to each directly (see next).

\textbf{Step 3:} the codechecker runs the code, based on instructions provided by
the author, and checks if some/all of the results from the paper can be
reproduced. (See later for exact nature of ``reproduced''). If there are
any problems running the code, the codechecker asks the author for help
to resolve the problems. The codechecker then tries to run the code
again. This process iterates until either the codechecker is successful,
or the codechecker concludes the code does not (\emph{what do in this
case}). As part of this process, the codechecker could work entirely
locally on their own compute resource, or in the cloud, e.g.~using
mybinder infrastructure. Using such cloud-based infrastructure allows
for the codechecker and author to collaboratively run the code together.

\textbf{Step 4:} the codechecker writes a certificate stating how the code was
run and includes a copy of outputs (figures, tables) that were
independently generated.

\textbf{Step 5:} the certificate and code/data get deposited on an open archive
(currently Zenodo), and the certificate is given to the PUBLISHER.

\textbf{Step 6:} the PUBLISHER can then use the certificate as part of their
supplementary records for a paper, and give credit to the codechecker
for their work by depositing appropriate records (\emph{bit hazy here}).

\begin{figure}
  \centering
      \includegraphics{figs/codecheck_overview.pdf}
  \caption{The detective. Could we number the arrows as steps?
(Implementing the CODECHECK process).  DANIEL: could you check whether
the icons are reusable.}
  \label{fig:worfklow}
\end{figure}

@@

DANIEL: COULD YOU DO THIS?
This example workflow leaves sufficient room for different
implementations of this process. There are several aspects that can be
considered:

@@

We have a curated list of 15 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons (do we name them all?)
from interesting historical papers, to papers codechecked as part of
peer review, or because of public interest (covid). several covid models
have been done during the time of peer review of these papers, under our
initative rather than requested from journal, but certificate have then
been acknowledged/cited in the paper (LANCET / LSHTM).
Table~\ref{tab:register} lists the certificates completed to date,
along with citations to the original papers.

@@

\begin{table}
  \centering

  \begin{tabular}{lllp{9cm}}  \toprule
    Certificate & Ref & Research area & Description \\ \midrule
    2020-001 & \cite{cert-2020-001} & Machine learning & CODECHECK undertaken post acceptance of manuscript and before its publication in \textit{Gigascience} \cite{Piccolo2020-lo}.
    \\ 
    2020-002 & \cite{cert-2020-002} & neuroscience &
    Paper from 1997 showing unsupervised learning from natural images \cite{Hancock1992-mp}.
    \\
    2020-003 & \cite{cert-2020-003} & neuroscience &
    Classic paper from 1982 on  models of associative memory \cite{Hopfield1982-mz}. \\
    2020-004 & \cite{cert-2020-004} & area & barto \\
    2020-005 & \cite{cert-2020-005} & area & Larisch \\
    2020-006 & \cite{cert-2020-006} & area & Detorakis \\
    2020-007 & \cite{cert-2020-007} & area & in progress \\
    2020-008 & \cite{cert-2020-008} & Covid & covid-uk Davies \\
    2020-009 & \cite{cert-2020-009} & Covid &  tracing \\
    2020-010 & \cite{cert-2020-010} & Covid &  report 9 \\
    2020-011 & \cite{cert-2020-011} & Covid &  nature europe \\
    2020-012 & \cite{cert-2020-012} & Covid &  USA Covid modelling \cite{Unwin2020}. \\
    \\ \bottomrule
  \end{tabular}
  \caption{Register of completed certificates.  An interactive version
  of this is available at \url{http://codecheck.org.uk/register}.}
  \label{tab:register}
\end{table}

@@

At the completion of each CODECHECK, a certificate is written that
states which figures and tables from the original article could be
reproduced by the CODECHECKER.  This certificate is made freely
available so that all readers of the article can see which elements
have been reproduced.  The certificate also includes links to the code
and data used by the codechecker, allowing others to build on the
work.  The format of the certificates has evolved slightly during the
project, as we have learnt to automate different aspects of the
certification.

To demonstrate the typical contents of a certificate, in
Figure~\ref{fig:cert} we have annotated the first four (out of ten)
pages of the certificate 2020-012 \cite{cert-2020-012}.  The reference
article \cite{Unwin2020} describes the modelling COVID-19 across the
USA.
%% TODO - has this been published yet?.  No, not yet... in review...
Figure~\ref{fig:cert}A shows the certificate number, and its DOI,
which points to the certificate as archived on Zenodo.  Our logo is
attached to the certificate to denote successful reproduction.
Figure~\ref{fig:cert}B provides the key metadata about the check --
which paper was checked, and who/when it was checked.  Materials
relating to the codecheck are freely available via the Repository
link.  Figure~\ref{fig:cert}C is a summary of how the codecheck was
performed, and any interesting findings.

Figure~\ref{fig:cert}D (page 2 of the certificate) shows a table that
lists the outputs that were generated from the codecheck (this is
termed the MANIFEST.)  For each output, there is a comment stating
which figure/table it should be compared to in the original paper,
along with the size of the file.  On page 3 of the certificate,
Figure~\ref{fig:cert}E shows some more detailed notes from the
codechecker, in this case indicating what steps were needed to
initialize the environment and run the code.  This particular example
was quite long, taking about 17 hours to run.  Finally, on page 4 of
the certificate, we see the first output that was generated by the
codecheck (Figure~\ref{fig:cert}F).  In this case, the figure matches
figure 4 of REF.  Subsequent pages of the certificate show other
outputs (not shown here).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/annotate-cert-crop.pdf}
  \caption{Annotated certificate 2020-012 (first four pages only).}
  \label{fig:cert}
\end{figure}

\subsubsection*{Tools and resources}\label{tools}

To run the system, we rely on github to host our work, and use a
custom R pacakge, codecheck, to help with the generation of
certificates and depositing work on zenodo. (Cite zen4R).  You can do
everything with open infrastructure (e.g.~automated rendering of
register). but you can use your own tools if you prefer.

@@

\textbf{Checking multiple times?}  If you check a preprint before peer
review, it might need checking again as it is modified during peer
review.  Unavoidable, and as happened to us (2020-012 has been
revised).  But checking the manuscript the second time is likely to be
much less work than the first time.

As a principle, we do not require results to be exactly the same for a
codecheck to pass, simply that the code runs and generates the output
files that the author claims. Stochastic simulations mean that often
we cannot get exactly the same, or even different versions of
libraries.  (Cite freesurfer example \cite{Gronenschild2012-pp} that
generates different outputs on different operating systems).

\begin{itemize}
\item modelDB principle: minimum of one figure should be reproducibe.
  Example certificate 2020-016 (I think) which had only a small subset
  of the possible figures. Still useful and worth sharing.
\end{itemize}
